Local wind regime induced by the presence of giant linear dunes.

### 2021-12-31 Paper/main.tex
Rewieving all paper + appendices

\title{Significant wind disturbances induced by giant dunes.}

@@

\begin{abstract}
  abstract
\keywords{Boundary layer \and Turbulent flow \and Sand dunes \and Fluide-structures interactions}
\end{abstract}

@@

Whenever a flow encounters an obstacle, different interactions can arise depending on the different time and length scales involved. In the case of atmospheric flows, this mainly depends on the part of the atmosphere, schematically composed of a turbulent boundary layer topped by a part free of turbulence, with which the obstacle interacts \citep{Stull1988}.
%
At the largest scale, the feedback of mountains on the stratified flow of the free atmosphere results in wave generation as well as significant wind disturbances, such as downslope winds in the lee side \citep{Duran1990}. Inside the boundary layer, the interaction between a turbulent flow and hilly surfaces is for example key to the understanding ocean surface wind-driven waves, or eolian bedforms in desert \citep{Belcher1998, Sullivan2010, Courrech2015}.

Indeed, eolian sand dunes typically emerge from the feedback of the topography on the turbulent flow, which speeds up close to the dune crest \citep{Rubin1987, Charru2013, Courrech2014}. Later on, when the dune reaches an intermediate size, it may also induce significant wind deflections. For exemple, this can affect the sediment pathways of coastal systems \citep{Hesp2015}, or the collective behavior of dune populations, through long-range interactions due to flow disturbances induced by each individual \citep{Smith2017, Bacik2020}. As the dunes increase in size by collisions and coarsening, they sometimes reach a giant size comparable to the boundary layer depth~\citep{andreotti2009}. However, the wind disturbances induced by these giant dunes have never been quantified.

The study of the feedback of obstacles on atmospheric flows allows its incorporation into numerical meteorological models. The latter then become mainly limited by the accuracy of the included topographical data, as well as by the spatial grid of the model. For example, the ERA5-Land climate reanalysis is limited by its $9~\textup{km}$ spatial resolution, while including the data $30$-m Digital Elevation Models (DEMs) of the shuttle radar topography mission \citep{Farr2007, munoz2021}. As such, it can not reproduce the flow disturbances induced by giant dunes, which have a typical length scale $\sim 1~\textup{km}$.

Here, we compare the wind speed and direction of the ERA5-Land dataset to local measurements in four different places across the Namib desert. Where no significant topographies smaller than the model grid are present, we show that the two wind datasets agree with each other. On the contrary, in places with giant dunes, we show that they may differ for some specific meteorological conditions, that we link to the circadian cycle of the atmospheric boundary layer. In doing so, we highlight the importance of medium-scale topographies for local wind regimes and, in the case of sand seas, draw implications for smaller-scale eolian bedforms.

@@

In this study, we focus on four places across and nearby the Namib desert, highlighting different environments (see Fig.~\ref{Fig1}).
%
The Adamax station is located near the Adamax salt pan, in a highly vegetated area. The Huab station, located on the coast at the outlet of the Huab river is an arid environment exhibiting $60$-m scale barchan dunes. While these two stations are in environments with no mid-scale topography, this is not the case for the Deep Sea and South Namib stations. Both are located in the interdune between giant linear dunes with kilometric wavelengths and superimposed patterns. In this section, we describe and compare the wind regimes resulting from the available datasets in each station.

  \begin{figure}
    \centering
    \includegraphics[scale=1]{Figures/Figure1.pdf}
    \caption{Wind data used in this study \textbf{a}: Location of the studied sites. \textbf{b--e}: Satellite images of the studied sites (Google-Earth, Maxar Technologies, CNES/Airbus). In each subplot, the left and right wind roses represent the data from the ERA5Land climate reanalysis and the local wind stations, respectively. Note that the bars show the direction towards which the wind blows. The red dots show the location of local wind stations.}
    \label{Fig1}
  \end{figure}

@@

  Two wind datasets are used in this study. First, local winds are provided by stations situated in the four different places (see Fig.~\ref{Fig1}). The wind strength and direction are measured every 10 minutes by cup anemometers and wind vanes, at heights between $2$~m and $3$~m depending on the station. The available period of measurements ranges from 1 to 5 discontinuous years distributed between 2012 and 2020 (see Fig.~\ref{Fig1_supp}). We checked that at least one complete seasonal cycle is available at each station.
  %
  Then, regional winds are extracted at the same locations and periods from the ERA5-Land dataset, which is a replay at a smaller spatial resolution of ERA5, the latest climate reanalysis from the ECMWWF \citep{Hersbach2020, munoz2021}. It provides hourly estimates of the 10-m wind velocity and direction at a spatial resolution of $\sim 9~\textup{km}$ ($0.1^\circ\times0.1^\circ$).

  For comparison, the local measurements are averaged into 1-hr bins centered on the temporal scale of the ERA5-Land estimates (see Fig.~\ref{Fig2_supp}). As the wind velocities of both datasets are provided at different heights, we convert them into shear velocities (see SI section 1), characteristic of the whole turbulent wind profile within the atmospheric boundary layer, which are then used together with the wind direction for further analysis. The resulting wind data are shown on the wind roses of Fig.~\ref{Fig1}(b--e).

@@

  The obtained wind regimes are shown in figure~\ref{Fig1}. In the Namib, the regional wind patterns are essentially controlled by the see breeze, resulting in strong northward components (sometimes slightly deviated by the large scale topography) present in all regional wind roses \citep{lancaster1985}. These daily winds are dominant during the second-half of the year (Septembre-January). In winter, an additional easterly component can be recorded during the night, induced by the combination of katabatic winds forming on the mountains, and infrequent `berg' winds, which are responsible of the high wind velocities observed \citep{lancaster1984}. The frequency of these easterly components decreases from the inland to the coast, resulting in bidirectional wind regimes within the Namib Sand Sea and at the Adamax salt pan (Fig.~\ref{Fig1}b, \ref{Fig1}d and \ref{Fig1}e) and a unidirectional wind regime on the coast at the outlet of the Huab River (Fig.~\ref{Fig1}c).

  In the case of the Adamax and Huab stations, the regional wind roses qualitatively match those corresponding to the local in situ measurements. However, for the Deep Sea and South Namib stations, the local wind roses exhibit additional components aligned with the giant dune orientation visible on the satellite images (Fig.~\ref{Fig1}c--d).
  %
  Indeed, the analysis of the wind speed and direction time series shows that the agreement between the local and regional datasets is always verified when no mid-scale topography are present (Fig.~\ref{Fig2}a--b) and Fig.~\ref{Fig5_supp}). In contrast, for the stations within the giant dune field, we observe that this agreement is limited to the Septembre--January time periods (Fig.~\ref{Fig2}c--d).

@@

  When giant dunes are present, in the February--August period, the local and regional winds match only during the morning, i.e when the southerly/southwesterly sea breeze dominates (see Fig.~\ref{Fig2}(e--f), Fig.~\ref{Fig3} and Fig.~\ref{Fig6_supp}). In the late afternoon and during the night, when the northwesterly `berg' and katabatic winds blow, the two datasets differ. In this case, the angular wind distribution of the local measurements exhibits two additional modes separated of $\simeq 180^\circ$, each corresponding to the giant dune alignement (purple frame in Fig.~\ref{Fig3} and Fig.~\ref{Fig6_supp}, as well as Fig.~\ref{Fig7_supp}). This deviation is also associated with a global attenuation of the wind strength (Fig.~\ref{Fig8_supp}). Remarkably, all these figures show that this process occurs for low wind velocities, typically for $u_{*} < 0.1~\textrm{m}~\textrm{s}^{-1}$. For shear velocities larger than $0.25~\textrm{m}~\textrm{s}^{-1}$, this wind reorientation does not occur. Finally, for intermediate shear velocities, both reorientation along the dune crest and no reorientation are observed (Fig.~\ref{Fig7_supp}).

@@

  In the case of linear ridges, dune-induced flow disturbances have mainly been related to the incident wind direction~\citep{Walker2009, Hesp2015}. In our case, it is unlikely to be the dominant parameter, as the most deflected wind for both stations is the most perpendicular, where it should be winds with incident directions between $30^{\circ}$ and $70^{\circ}$~\citep{Hesp2015}. An important observation is the difference in behavior between low and high wind velocities, which suggests a change in the hydrodynamical regime.

  Previous studies have linked atmospheric flow around and over topographical obstacles to the vertical structure of the atmosphere \citep{Stull1988}. More particularly, dunes evolves in its lower part, the turbulent atmospheric boundary layer (ABL), typically characterized by a logarithmic wind profile and a vertically constant potential temperature. Above, the free atmosphere (FA) is a stably stratified zone in which turbulence is negligible, and where the flow is usually considered as incompressible and inviscid. In the middle, a transitional layer, also known as entrainment zone, is characterized by a sharp increase of the potential temperature, which traps the turbulence resulting from the surface friction below it.

  In the following, we sum-up the dominant numbers leading to different hydrodynamical interactions with topographical obstacles, and interpret the data with respect to the corresponding physical mechanisms.

  \begin{figure}
    \centering
    \includegraphics[scale=1]{Figures/Figure4.pdf}
    \caption{\textbf{a--b}: Vertical profiles of the virtual potential temperature at 2 different time steps (blue - 29/11/2012 - 1100 UTC, red - 21/03/2017 - 1200 UTC) at the Deep Sea station. Dots: data from the ERA5 reanalysis. Transparent dashed lines: boundary layer height given by the ERA5 reanalysis, calculated from the bulk Richardson number \citep{seidel2012}. Plain lines: vertical (boundary layer) and linear (free atmosphere) fits to estimate the stratification properties. \textbf{c--f}: Sketches representing the interaction between the giant dunes and the atmospheric flow for different meteorological conditions. \textbf{g}: Streamlines qualitatively representing the effect of low, medium and high flow confinement. For details on the streamline derivation, see Appendix~\ref{turbulent_wind_model}.}
    \label{Fig4}
  \end{figure}

  \subsection{Relevant non-dimensional parameters and physical modeling}
  \label{theoretical_framework}

  Flow deflection over ridges can be simplistically understood from a balance between inertia and pressure gradients~\citep{Hesp2015}. As the flow approaches the ridge crest, the compression of the streamlines results in larger flow velocities, and thus lower pressures~\citep{Rubin1987}. An incident flow oblique to the ridge is then deflected towards lower pressure zones, i.e towards the crest. Turbulent dissipation at the bottom and non-linearities tends to increase this effect downstream, resulting in along the crest wind deflection in the lee side~\citep{Hesp2015, Gadal2019}.

  Another way to increase the flow deflection is its confinement below a capping surface, that results in further streamline compression. This happens when the flow disturbance induced by the obstacle reaches the surface. As obstacles typically disturb flow over a characteristic height similar to their width, the potential of interaction between the dunes and the overlying atmospheric structure is well captured by the parameter $k H$, where $k = 2\pi/\lambda$ is the wavenumber and $H$ the ABL depth. Here, the giant dunes have kilometric wavelengths, such that $0.02 \lesssim k H \lesssim 5$, and they interact most of the time with the capping layer and the stratified free atmosphere above \citep{andreotti2009}.

  Note that the ability of the capping layer and stratification to accommodate a perturbation induced by the topography directly impacts the strength of this confinement effect (Fig.~\ref{Fig4}). This is typically quantified using surface and internal Froude numbers
  \citep{Vosper2004, Stull2006, Sheridan2006, Hunt2006, Jiang2014}:
  \begin{equation}
        \mathcal{F}r_{\textup{S}} = \displaystyle\frac{U}{\sqrt{\displaystyle\frac{\Delta\rho}{\rho}gH}}, \, \mathcal{F}r_{\textup{I}} = \displaystyle\frac{k U}{N},
  \end{equation}
  where $U$ is the wind velocity at the top of the ABL, $\rho$ its average density, $\Delta\rho$ the density jump between the ABL and the FA and $N$ is the Brunt-Väisälä frequency, characteristic of the stratification.

  The smallest wind disturbances are expected during the day, when the ABL depth is comparable to the dune wavelength ($k H \gtrsim 1$) and for large wind velocities, which correspond to a weak confinement situation (Fig.~\ref{Fig4}d). On the contrary, large wind disturbances are expected to occur during the night, when the confinement is mainly induced by shallow ABL (Fig.~\ref{Fig4}e--f). Note that this strong confinement can be somewhat reduced in the case of strong winds (corresponding to large Froude numbers, see Fig.~\ref{Fig4}f), explaining the threshold in velocity observed in the data (see section~\ref{section_data_feedback}).

@@

  To highlight these different regimes from our data, we compute wind disturbance diagrams in the space defined by the three relevant non-dimensional numbers presented above, $\left(k H,\, \mathcal{F}r_{\textup{S}}, \, \mathcal{F}r_{\textup{I}}\right)$. Those are calculated from the time series of the geopotential, temperature and specific humidity vertical profiles available in the ERA5 climate reanalysis (see SI section 2).
  %
  Flow deviation is computed as the minimal angle between the wind orientations from the two datasets:
  \begin{equation}
    \delta_{\theta} = \left\vert \textup{min}\left(\left[\theta_{\textup{ERA}} - \theta_{\textup{station}}\right] \, \textup{mod.} \, 360, \left[\theta_{\textup{station}} - \theta_{\textup{ERA}}\right] \, \textup{mod.} \, 360\right) \right\vert.
  \end{equation}
  %
  The relative velocity modulation is computed as
  %
  \begin{equation}
    \delta_{\textup{u}} = \frac{u_{*}^{\textup{ERA}} -  u_{*}^{\textup{station}}}{u_{*}^{\textup{ERA}}}.
  \end{equation}

  As both Froude numbers have qualitatively the same impact on the flow confinement, we first leave aside the internal Froude number, and focus on the space $(k H,\, \mathcal{F}r_{\textup{S}})$. When representing the two variables $\delta_{\theta}$ and $\delta_{\textup{u}}$ in this space, different regime emerges (Fig.~\ref{Fig5}). Small wind disturbances ($\delta_{\theta} \to 0$, $\delta_{\textup{u}} \to 0$) are located in the top-right part of the diagrams, corresponding to a regime mixing low-interaction and low-confinement ($k H$ and $\mathcal{F}r_{\textup{S}}$ large enough, Fig.~\ref{Fig4}d).
  %
  Lower values of $k H$ (stronger interaction) or Froude numbers (stronger confinement) then both leads to an increase in wind disturbances, both in terms of orientation and velocity.   Below a threshold value of $k H \simeq 0.3$, wind disturbance occurs independently of the Froude numbers value. Furthermore, this also seems to control a transition between from damped to amplified wind velocities within the interdune (Fig.~\ref{Fig5}b), for which we do not have an explanation. Note that the same interpretation can be done with the diagrams including the internal Froude number $\mathcal{F}r_{\textup{I}}$, as shown by Fig.~\ref{Fig12_supp}.

  Interestingly, the limit of no-interactions between the topography and the boundary layer structure ($k H \gg 1$), in which the properties of the capping layer and the stratification become irrelevant, is never reached here, in the case of giant dunes.

  \begin{figure}
    \centering
    \includegraphics[scale=1]{Figures/Figure5.pdf}
    \caption{Regime diagrams of the wind deviation $\delta_{\theta}$ and relative attenuation/amplification $\delta_{u}$ in the space $(\mathcal{F}r_{\textup{S}}, \, kH)$, containing the data from both the Deep Sea and South Namib stations. Blue dashed lines empirically delimit the different regimes. The point density in each bin of the diagrams is shown in Fig.~\ref{Fig11_supp}. The regime diagrams in the spaces $(\mathcal{F}r_{\textup{I}}, \, kH)$ and $(\mathcal{F}r_{\textup{I}}, \, \mathcal{F}r_{\textup{S}})$ are shown in Fig.~\ref{Fig12_supp}.}
    \label{Fig5}
  \end{figure}

@@

 The comparison of local (direct measurements) and regional (climate reanalysis) wind data reveals the giant dunes feedback on the flow. In flat areas, the matching between both datasets confirms the ability of the ERA5Land climate reanalysis to predict the wind flow down to scales $\sim 10~\textrm{km}$, i.e the grid model. When smaller scale topographies are present (giant dunes in our case), locally measured wind regimes may significantly differ from the regional ones. Furthermore, we link these disturbances induced by the dunes to their interaction with the lower part of the atmospheric vertical structure, and more specifically to its circadian variability. During the night, the presence of a shallow atmospheric boundary layer (ABL) induces a strong flow confinement, associated with large wind deviation and acceleration or deceleration. During the day, the capping layer is high enough to prevent its interaction with the giant dunes, resulting in a low flow confinement, and thus smaller wind disturbances. Interestingly, we also found that this effect could be counterbalanced by the presence of large wind velocities, capable of deforming the capping layer and/or the FA stratification, thus decreasing the confinement effect.

>>>>

\title{Local wind regime induced by the presence of giant linear dunes.}

@@

\begin{abstract}

  \keywords{Boundary layer \and Turbulent flow \and Sand dunes \and Fluide-structures interactions}
\end{abstract}

@@

As a flow encounters an obstacle, different interactions can arise depending on the different time and length scales involved. In the case of atmospheric flows, this mainly depends on the part of the atmosphere, schematically composed of a turbulent boundary layer topped by a stably stratified one, with which the obstacle interacts \citep{Stull1988}.
%
At the largest scale, the feedback of mountains on the stratified part results in wave generation as well as significant wind disturbances, such as downslope winds in the lee side \citep{Duran1990}. Inside the boundary layer, the interaction between a turbulent flow and hilly surfaces is for example key to the understanding ocean surface wind-driven waves~\citep{Sullivan2010}, or eolian bedforms in desert \citep{Courrech2015}.

Looking at the flow close to the surface, two different components of the topography feedback can be isolated. First, the flow accelerates on the upwind slope, and slows down on the downwind one, with a maximum velocity slightly upwind of the crest~\citep{Jackson1975, Sykes1980, Hunt1988}. Several measurements are available, performed in subaqueous flumes~\citep{Zilker1977, Zilker1979, Frederick1988} and in eolian field conditions~\citep{Claudin2013, Lu2021}. Second, deflection occurs when the incident flow direction is not perpendicular to the ridge crest. While predicted to be small (less than $10^{\circ}$) in the linear regime~\citep{Gadal2019}, significant flow steering has been reported in the field on the downwind side of larger aspect ratios obstacles, such as mountain ranges~\citep{Kim2000, Lewis2008, Fernando2019} or sand dunes~\citep{Walker2009, Walker2013, Hesp2015, Smith2017}.

In the case of aeolian bedforms, the measurements mentioned above were made on dunes a few meters high only. The resulting wind disturbances have then been shown to significantly impact the sediment pathways of the system \citep{Hesp2015}, but also to affect the collective behaviour of dune populations trough long-range interactions \citep{Bacik2020}.
%
However, eolian dunes coexists in sand seas at different scales, each representative of their different evolution stages, from emergence at twenty meters wavelengths to giant size, corresponding to kilometric wavelengths and heights of several dozen meters \citep{McKee1979}. Quantifying these two effects for giant dunes is then all the more important for several reasons.
%
First, larger wind disturbances are expected from larger obstacles.
%
Second, the flow--structure coupling is not straightforward due to vertical structure of the atmosphere with which the giant dunes interact due to their sizes \citep{andreotti2009}.
%
Finally, the morphodynamics of aeolian bedforms are strongly dependent on the wind regime~\citep{Livingstone1996}. Thus, small-scale bedforms are sensitive to wind disturbances induced by larger-scale bedforms. This has been demonstrated for impact ripples on dunes \citep{Howard1977,Hood2021}, but has never been studied for medium-scale dunes near giant dunes.

In the literature, arid areas have been studied at the desert scale from climate reanalyses based on global atmospheric models \citep{Blumberg1996, Livingstone2010, Ashkenazy2012, Jolivet2021, Hu2021}, such as ERA-40, ERA-Interim or ERA-5 \citep{Uppala2005, Dee2011, Hersbach2020}. However, the spatial resolution (tens of kilometers) of these reanalyses implies average quantities that do not resolve the smaller scales, ranging from the study of individual dunes to the border of mountains \citep{Livingstone2010}. Lately, the release of ERA5-Land allow to push back this limitation by providing up to 70 years of hourly wind predictions at a 9~km spatial resolution \citep{munoz2021}. However, due to its recent nature, the applicability limitations of this dataset remain to be studied.

Here, we compare local wind speeds and directions measured in four different places across the Namib desert to the regional predictions of the ERA5-Land climate reanalysis. When the measurement stations are surrounded by a relatively flat environment, we show that local measurements and regional predictions agree with each other. On the contrary, in the sand sea, we show that they differ due to the presence of giant dunes. Furthermore, we link the magnitude of these discrepancies to the circadian cycle of the atmospheric boundary layer. Finally, we draw implications for smaller-scale eolian bedforms.

@@

  We focus on four places across and nearby the Namib desert, highlighting different environments (see Fig.~\ref{Fig1}).
  %
  The Adamax station is located near the Adamax salt pan, in a highly vegetated area. The Huab station, on the coast at the outlet of the Huab river, is in an arid environment exhibiting $60$-m scale barchan dunes. While the surroundings of these two stations are relatively flat, this is not the case for the Deep Sea and South Namib stations. Both are in the interdune between tens of meters high giant linear dunes with kilometric wavelengths and superimposed patterns. In this section, we describe and compare winds from local measurements and climate reanalysis predictions.

  \begin{figure}
    \centering
    \includegraphics[scale=1]{Figures/Figure1.pdf}
    \caption{Wind data used in this study \textbf{a}: Location of the studied sites. \textbf{b--e}: Satellite images of the studied sites (Google-Earth, Maxar Technologies, CNES/Airbus). In each subplot, the left and right wind roses represent the data from the ERA5Land climate reanalysis and the local wind stations, respectively. Note that the bars show the direction towards which the wind blows. The black dots show the location of local wind stations.}
    \label{Fig1}
  \end{figure}

@@

  Local winds are provided by measurement stations located in the four different places (see black dots in Fig.~\ref{Fig1}). The wind strength and direction are sampled every 10 minutes by cup anemometers and wind vanes, at heights between $2$~m and $3$~m depending on the station. The available period of measurements ranges from 1 to 5 discontinuous years distributed between 2012 and 2020 (see Fig.~\ref{Fig1_supp}). We checked that at least one complete seasonal cycle is available at each station.
  %
  Regional winds are extracted at the same locations and periods from the ERA5-Land dataset, which is a replay at a smaller spatial resolution of ERA5, the latest climate reanalysis from the ECMWWF \citep{Hersbach2020, munoz2021}. It provides hourly predictions of the 10-m wind velocity and direction at a spatial resolution of $\sim 9~\textup{km}$ ($0.1^\circ\times0.1^\circ$).

  For comparison, the local measurements are averaged into 1-hr bins centered on the temporal scale of the ERA5-Land estimates (see Fig.~\ref{Fig2_supp}). As the wind velocities of both datasets are provided at different heights, we convert them into shear velocities (see SI section 1), characteristic of the turbulent wind profile, which are then used together with the wind direction for further analysis. The resulting wind data are shown on the wind roses of Fig.~\ref{Fig1}(b--e).

@@

  The obtained wind regimes are shown in figure~\ref{Fig1}. In the Namib, the regional wind patterns are essentially controlled by the see breeze, resulting in strong northward components (sometimes slightly deviated by the large scale topography) present in all regional wind roses \citep{lancaster1985}. These daytime winds are dominant during the second-half of the year (Septembre-January). In winter, an additional easterly component can be recorded during the night, induced by the combination of katabatic winds forming on the mountains, and infrequent `berg' winds, which are responsible of the high wind velocities observed \citep{lancaster1984}. The frequency of these easterly components decreases from the inland to the coast, resulting in bidirectional wind regimes within the Namib Sand Sea and at the Adamax salt pan (Fig.~\ref{Fig1}b, \ref{Fig1}d and \ref{Fig1}e) and a unidirectional wind regime on the coast at the outlet of the Huab River (Fig.~\ref{Fig1}c).

  In the case of the Adamax and Huab stations, the wind roses from the regional predictions qualitatively match those corresponding to the local measurements. However, for the Deep Sea and South Namib stations, the measured wind roses exhibit additional components aligned with the giant dune orientation visible on the satellite images (Fig.~\ref{Fig1}c--d).
  %
  The time series of wind speed and direction show that this agreement in the case of Adamax and Huab stations is always verified (Fig.~\ref{Fig2}a--b) and Fig.~\ref{Fig5_supp}). In contrast, for the stations within the giant dune field, we observe that this agreement is limited to Septembre--January time periods (Fig.~\ref{Fig2}c--d).

@@

  In the February--August period, when giant dunes are present, the local and regional winds match during daytime only, i.e when the southerly/southwesterly sea breeze dominates (see Fig.~\ref{Fig2}(e--f), Fig.~\ref{Fig3} and Fig.~\ref{Fig6_supp}). In the late afternoon and during the night, when the northwesterly `berg' and katabatic winds blow, measurements and predictions differ. In this case, the angular wind distribution of the local measurements exhibits two additional modes separated of $\simeq 180^\circ$, each corresponding to the giant dune alignement (see the purple frame in Fig.~\ref{Fig3} and Fig.~\ref{Fig6_supp}, as well as Fig.~\ref{Fig7_supp}). This deviation is also associated with a global attenuation of the wind strength (Fig.~\ref{Fig8_supp}). Remarkably, all these figures show that this process occurs for low wind velocities, typically for $u_{*} < 0.1~\textrm{m}~\textrm{s}^{-1}$. For shear velocities larger than $0.25~\textrm{m}~\textrm{s}^{-1}$, this wind reorientation does not occur. Finally, for intermediate shear velocities, both reorientation along the dune crest and no reorientation are observed (Fig.~\ref{Fig7_supp}).

@@

  For linear ridges, dune-induced flow disturbances have mainly been related to the angle between wind direction and crest orientation, with a maximum for angles between $30^{\circ}$ and $70^{\circ}$~\citep{Walker2009, Hesp2015}. In our case, the most deflected wind for both stations is the most perpendicular, such that the incident wind direction does not seem to be the dominant parameter controlling the wind deflection. In contrast, a different behavior is observed between  low and high wind velocities, suggesting a change in hydrodynamical regime.

  In the following, we discuss the relevant parameters leading to different hydrodynamical interactions with topographical obstacles, and interpret the data with respect to the corresponding physical mechanisms.

  \begin{figure}
    \centering
    \includegraphics[scale=1]{Figures/Figure4.pdf}
    \caption{\textbf{a--b}: Vertical profiles of the virtual potential temperature at 2 different time steps (day - 31/03/2017 - 1200 UTC, night - 21/03/2017 - 2200 UTC) at the Deep Sea station. Dots: data from the ERA5 reanalysis. Transparent dashed lines: boundary layer height given by the ERA5 reanalysis, calculated from the bulk Richardson number \citep{seidel2012}. Plain lines: vertical (boundary layer) and linear (free atmosphere) fits to estimate the stratification properties. \textbf{c--f}: Sketches representing the interaction between the giant dunes and the atmospheric flow for different meteorological conditions. \textbf{g}: Streamlines qualitatively representing the effect of low, medium and high flow confinement. For details on the streamline derivation, see Appendix~\ref{turbulent_wind_model}.}
    \label{Fig4}
  \end{figure}

  \subsection{Relevant non-dimensional parameters and physical considerations}
  \label{theoretical_framework}

  Flow deflection over ridges can be understood from the Bernoulli principle~\citep{Hesp2015}. As the flow approaches the ridge crest, the compression of the streamlines results in larger flow velocities, and thus lower pressures~\citep{Rubin1987}. An incident flow oblique to the ridge is then deflected towards lower pressure zones, i.e towards the crest. Turbulent dissipation at the bottom and non-linearities tends to increase this effect downstream, resulting in along the crest wind deflection in the lee side~\citep{Hesp2015, Gadal2019}.

  Another way to increase the flow deflection is its confinement below a capping surface, that results in further streamline compression. This is the case for bedforms forming in open channel flows such as rivers~\citep{Fourriere2010, Unsworth2018}, but also for eolian dunes. These dunes evolve in the turbulent atmospheric boundary layer (ABL), which is capped by a transitional layer separating it from the stratified atmosphere above (see Fig.~\ref{Fig4}). Two different mechanisms control the possibility of this additional streamline compression.

  On one hand, it depends if the flow disturbance induced by the underlying topography reach the surface. As obstacles typically disturb flow over a characteristic height similar to their length, the potential of interaction between the dunes and the overlying surface is well captured by the parameter $k H$, where $k = 2\pi/\lambda$ is the wavenumber and $H$ the ABL depth. Note that $H$ is directly related to the radiative fluxes at the Earth surface, and thus varies with the circadian and seasonal cycles. Here, the giant dunes have kilometric wavelengths, such that $0.02 \lesssim k H \lesssim 5$, and they interact most of the time with the capping layer and the stratified free atmosphere (FA) above \citep{andreotti2009}. Interestingly, the limit of no-interactions between the topography and the boundary layer structure ($k H \gg 1$), in which the properties of the overlying atmospheric structure are irrelevant, is never reached here, in the case of giant dunes.

  On the other hand, it depends on rigidity of the capping surface, as its deformation releases the confinement effect inducing streamline compression. This is typically quantified using the Froude number~\citep{Vosper2004, Stull2006, Sheridan2006, Hunt2006, Jiang2014}:
  \begin{equation}
        \mathcal{F} = \displaystyle\frac{U}{\sqrt{\displaystyle\frac{\Delta\rho}{\rho}gH}},
  \end{equation}
  where $U$ is the wind velocity at the top of the ABL, $\rho$ its average density, $\Delta\rho$ the density jump between the ABL and the FA.

  % Note that the ability of the capping layer and stratification to accommodate a perturbation induced by the topography directly impacts the strength of this confinement effect (Fig.~\ref{Fig4}). This is typically quantified using surface and internal Froude numbers
  % \citep{Vosper2004, Stull2006, Sheridan2006, Hunt2006, Jiang2014}:
  % \begin{equation}
  %       \mathcal{F} = \displaystyle\frac{U}{\sqrt{\displaystyle\frac{\Delta\rho}{\rho}gH}}, \, \mathcal{F}_{\textup{I}} = \displaystyle\frac{k U}{N},
  % \end{equation}
  % where $U$ is the wind velocity at the top of the ABL, $\rho$ its average density, $\Delta\rho$ the density jump between the ABL and the FA and $N$ is the Brunt-Väisälä frequency, characteristic of the stratification.

  The smallest wind disturbances are expected during the day, when the ABL depth is comparable to the dune wavelength ($k H \gtrsim 1$) and for large wind velocities, which correspond to a weak confinement situation (Fig.~\ref{Fig4}d). On the contrary, large wind disturbances are expected to occur during the night, when the confinement is mainly induced by shallow ABL (Fig.~\ref{Fig4}e--f). Note that this strong confinement can be somewhat reduced in the case of strong winds (corresponding to large Froude numbers, see Fig.~\ref{Fig4}f), explaining the transition from deflected to non-deflected winds related to low and high velocities observed in the data (see section~\ref{section_data_feedback}).

@@

  To highlight these different regimes from our data, we compute wind disturbance diagrams in the space defined by the two relevant non-dimensional numbers presented above, $\left(k H,\, \mathcal{F}\right)$. Those are calculated from the time series of the geopotential, temperature and specific humidity vertical profiles available in the ERA5 climate reanalysis (see SI section 2).
  %
  Flow deviation is computed as the minimal angle between the wind orientations from the local measurements, and the regional predictions. The relative velocity modulation is computed as
  %
  \begin{equation}
    \delta_{\textup{u}} = \frac{u_{*}^{\textup{ERA}} -  u_{*}^{\textup{station}}}{u_{*}^{\textup{ERA}}}.
  \end{equation}

  When representing the two variables $\delta_{\theta}$ and $\delta_{\textup{u}}$ in this space, different regime emerges (Fig.~\ref{Fig5}). Small wind disturbances ($\delta_{\theta} \to 0$, $\delta_{\textup{u}} \to 0$) are located in the top-right part of the diagrams, corresponding to a regime mixing low-interaction and low-confinement ($k H$ and $\mathcal{F}$ large enough, Fig.~\ref{Fig4}d).
  %
  Lower values of $k H$ (stronger interaction) or Froude number (stronger confinement) then both lead to an increase in wind disturbances, both in terms of orientation and velocity. Below a threshold value of $k H \simeq 0.3$, wind disturbance occurs independently of the Froude number value, probably due to enhanced non-linear effects linked to strong flow modulation by the obstacle in this part of the diagram. The Froude number also controls a transition from damped to amplified wind velocities in the interdune, with a transition at $\mathcal{F} \simeq 0.4$ (Fig.~\ref{Fig5}b). This may be linked to a transition in the flow regime in the lee side of the obstacle (lee waves, hydraulic jumps, rotors) but further measurements are needed in order to assess this \citep{baines1995, Vosper2004}.

  %
  % Furthermore, this also seems to control a transition between from damped to amplified wind velocities within the interdune (Fig.~\ref{Fig5}b), for which we do not have an explanation. Note that the same interpretation can be done with the diagrams including the internal Froude number $\mathcal{F}_{\textup{I}}$, as shown by Fig.~\ref{Fig12_supp}.

  \begin{figure}
    \centering
    \includegraphics[scale=1]{Figures/Figure5.pdf}
    \caption{Regime diagrams of the wind deviation $\delta_{\theta}$ and relative attenuation/amplification $\delta_{u}$ in the space $(\mathcal{F}, \, kH)$, containing the data from both the Deep Sea and South Namib stations. Blue dashed lines empirically delimit the different regimes. The point density in each bin of the diagrams is shown in Fig.~\ref{Fig11_supp}. The regime diagrams in the spaces $(\mathcal{F}_{\textup{I}}, \, kH)$ and $(\mathcal{F}_{\textup{I}}, \, \mathcal{F})$ are shown in Fig.~\ref{Fig12_supp}.}
    \label{Fig5}
  \end{figure}

  \subsection{On the influence of the stratification of the free atmosphere}

  The presence of a stratification in the free atmosphere can also impact the flow confinement, depending on its ability to deform under the presence of an underlying obstacle. This can be quantified using the internal Froude number~\citep{Vosper2004, Stull2006, Sheridan2006, Hunt2006, Jiang2014}:
  \begin{equation}
    \mathcal{F}_{\textup{I}} = \displaystyle\frac{k U}{N},
  \end{equation}
  where $N = \sqrt{-(g/\rho)(\partial \rho/\partial z)}$ is characteristic of the stratification. Both Froude numbers have the same qualitative effect on flow confinement, as they quantify the rigidity of the overlying layers. This is confirmed by figure~\ref{Fig12_supp}, where we can also find the different regimes related to wind disturbances described previously for the Froude number $\mathcal{F}$.

@@

  \begin{figure}
    \centering
    \includegraphics[scale=1]{Figures/Figure6.pdf}
    \caption{Implications for smaller scale patterns in (a) the South Namib and (b) Deep Sea. The ellipses indicates the diffe  rent types of elongating dunes, at large (plain) and small scale (dashed). The dune orientations are calculated using the model of \citet{Courrech2014} from the sand flux angular distributions, shown here for typical sand quartz grains of $180~\mu$m. The double blue and single red arrows correspond to the two possible dune growth mechanisms, bed instability and elongation, respectively. Likewise, plain arrows are calculated from the ERA5-Land data, and dashed arrows from the local measurements. Wedges show the uncertainty on the orientation calculation, and the arrows correspond to typical parameters found in the literature, i.e a grain diameter of $180~\mu$m and a flux-up ratio of 1.6. The black dots indicate the position of the measurement stations. See Appendix 2 for additional details.}
    \label{Fig6}
  \end{figure}

 The comparison of local (direct measurements) and regional (climate reanalysis) wind data reveals the giant dunes feedback on the flow. In flat areas, the matching between measurements and prediction confirms the ability of the ERA5Land climate reanalysis to predict the wind flow down to scales $\sim 10~\textrm{km}$, i.e the grid model. When smaller scale topographies are present (giant dunes in our case), locally measured wind regimes can significantly differ from the regionally predicted ones. Furthermore, we link these disturbances induced by the dunes to their interaction with the lower part of the atmospheric vertical structure, and more specifically to its circadian variability. During the night, the presence of a shallow atmospheric boundary layer (ABL) induces a strong confinement of the flow, associated with large wind deviation and acceleration or deceleration. During the day, the capping layer is high enough to prevent its interaction with the giant dunes, resulting in a low confinement of the flow, and thus smaller wind disturbances. Interestingly, we also found that this effect could be counterbalanced by the presence of large wind velocities, capable of deforming the capping layer and/or the FA stratification, thus decreasing the confinement effect.
